---
title: "Using gendertext"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using gendertext}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(gendertext)
```

## Introduction

The **gendertext** package provides simple, transparent tools for identifying gendered language in text and suggesting gender-neutral alternatives. 

It is designed for researchers, policy analysts, and practitioners who want to assess inclusive language practices in documents.

The package follows a dictionary-based approach, relying on a built-in corpus of gendered terms and suggested neutral replacements.

## Built-in dictionary

The package ships with a small example dictionary that is used internally by all functions.

```{r}
data("gender_dictionary")
head(gender_dictionary)

```

This dataset can be expanded or replaced in future versions of the package.

## Analysing text directly

The simplest way to use `gendertext` is to analyse a character string.

```{r}

gender_score(
  text = "Ladies and gentlemen, the chairman said he will call the policeman."
)

```

The reported neutral percentage is a proxy defined as tokens not matched to
any gendered term in the dictionary.


## Gender-neutral suggestions

```{r}
gender_suggestions(
  text = "Our chairman said he will email the mailman."
)

```

##Working with files

`gendertext` can also analyse files such as `.txt`, `.pdf`, and `.docx`
using the `readtext` package.

The package includes example files for demonstration purposes.


```{r}

txt <- system.file("extdata", "test.txt", package = "gendertext")
pdf <- system.file("extdata", "test.pdf", package = "gendertext")

gender_score(path = txt)
gender_suggestions(path = txt)

gender_score(path = pdf)

```

Please Note: PDF analysis depends on the presence of extractable text.
Scanned PDFs or image-only documents may not yield readable content.

##Limitations

- Results depend on dictionary coverage.

- The package does not attempt semantic interpretation.

- Gender neutrality is estimated using dictionary matching, not linguistic inference.


# Conclusion

`gendertext` offers a lightweight and reproducible way to examine gendered
language in text. Its transparent design makes it suitable for research,
policy review, and exploratory analysis.
